---
title: "2025 Day 3: Lobby"
format: html
execute:
  warning: false
---

## Part 1

### Prompt

You descend a short staircase, enter the surprisingly vast lobby, and are quickly cleared by the security checkpoint. When you get to the main elevators, however, you discover that each one has a red light above it: they're all offline.

"Sorry about that," an Elf apologizes as she tinkers with a nearby control panel. "Some kind of electrical surge seems to have fried them. I'll try to get them online soon."

You explain your need to get further underground. "Well, you could at least take the escalator down to the printing department, not that you'd get much further than that without the elevators working. That is, you could if the escalator weren't also offline."

"But, don't worry! It's not fried; it just needs power. Maybe you can get it running while I keep working on the elevators."

There are batteries nearby that can supply emergency power to the escalator for just such an occasion. The batteries are each labeled with their joltage rating, a value from 1 to 9. You make a note of their joltage ratings (your puzzle input). For example:

```
987654321111111
811111111111119
234234234234278
818181911112111
```

The batteries are arranged into banks; each line of digits in your input corresponds to a single bank of batteries. Within each bank, you need to turn on exactly two batteries; the joltage that the bank produces is equal to the number formed by the digits on the batteries you've turned on. For example, if you have a bank like 12345 and you turn on batteries 2 and 4, the bank would produce 24 jolts. (You cannot rearrange batteries.)

You'll need to find the largest possible joltage each bank can produce. In the above example:

In 987654321111111, you can make the largest joltage possible, 98, by turning on the first two batteries.
In 811111111111119, you can make the largest joltage possible by turning on the batteries labeled 8 and 9, producing 89 jolts.
In 234234234234278, you can make 78 by turning on the last two batteries (marked 7 and 8).
In 818181911112111, the largest joltage you can produce is 92.
The total output joltage is the sum of the maximum joltage from each bank, so in this example, the total output joltage is 98 + 89 + 78 + 92 = 357.

There are many batteries in front of you. Find the maximum joltage possible from each bank; what is the total output joltage?

### Solution

First, read in and parse the input file:

```{r}
read_input <- function(fname) {
  raw <- readLines(fname)
  split_strings <- strsplit(raw, "")
  lapply(split_strings, as.numeric)
}

test_input <- read_input("test_input.txt")
```

Our strategy will involve combinatorics -- we will derive every pair of digits in the vector in order, then take the maximum. We then apply that function over all elements of the input list.

```{r}
get_joltage <- function(bank) {
  # Get all combinations of 2 elements from `bank`
  combos <- combn(bank, 2)
  
  # Initialize a vector to store the results
  combo_vec <- numeric(ncol(combos))
  
  # Iterate over the columns of the combination matrix, combining the two rows
  # into a single numeric
  for (i in seq_len(ncol(combos))) {
    combo_vec[i] <- as.numeric(paste0(combos[1, i], combos[2, i]))
  }
  
  # Return the max value
  max(combo_vec)
}

get_total_joltage <- function(input) {
  sum(sapply(input, get_joltage))
}
```

Then glue it all together.

```{r}
solve_part_1 <- function(fname) {
  input <- read_input(fname)
  get_total_joltage(input)
}
```

Test data:

```{r}
solve_part_1("test_input.txt")
```

That matches the example in the prompt. Now for the real data:

```{r}
solve_part_1("input.txt")
```

And that's correct! On to part 2.
