---
title: "Advent of Code: 2022 Day 6 in R"
date: 2023-09-11
categories:
  - r
  - advent of code
  - puzzle
execute: 
  message: false
  warning: false
format: gfm
---

## Day 6: Tuning Trouble

## Part 1

This puzzle is a bit unique in that it give us 5 distinct test cases
- [test_input_1.txt](test_input_1.txt)
- [test_input_2.txt](test_input_2.txt)
- [test_input_3.txt](test_input_3.txt)
- [test_input_4.txt](test_input_4.txt)
- [test_input_5.txt](test_input_5.txt)

I think this will be fairly straightforward. We read in the data as a single long character string, then split it into a vector of individual characters. We iterate through that vector looking ahead at the next four characters. The first time that `length(unique(vec)) == 4`, we have found the first start-of-package marker.

```{r}
part_1 <- function(file) {
  data <- readLines(file)
  char_vector <- strsplit(data, "")[[1]]
  for (i in seq_along(char_vector)) {
    four_chars <- char_vector[i:(i + 3)]
    if (length(unique(four_chars)) == 4) {
      return(i + 3)
    }
  }
}
```

```{r}
part_1("test_input_1.txt") == 7
part_1("test_input_2.txt") == 5
part_1("test_input_3.txt") == 6
part_1("test_input_4.txt") == 10
part_1("test_input_5.txt") == 11
```

The examples look good. Now for the real data:

```{r}
part_1("input.txt")
```

Success!

## Part 2 

Now for part 2.

This should be the same, just switching out 14 for 4 when checking `length(unique(four_chars))`.

```{r}
part_2 <- function(file) {
  data <- readLines(file)
  char_vector <- strsplit(data, "")[[1]]
  for (i in seq_along(char_vector)) {
    four_chars <- char_vector[i:(i + 13)]
    if (length(unique(four_chars)) == 14) {
      return(i + 13)
    }
  }
}
```

```{r}
part_2("test_input_1.txt") == 19
part_2("test_input_2.txt") == 23
part_2("test_input_3.txt") == 23
part_2("test_input_4.txt") == 29
part_2("test_input_5.txt") == 26
```

All good. Now for the real data:

```{r}
part_2("input.txt")
```

Correct! This was a very easy challenge to do in R. It's possible that it would be more difficult in other languages. For example, in Go we would not have access to a function like `unique()` but would instead have to manually compare values or create a map with the characters in the vector to check for uniqueness. Regardless, this was straightforward to the point of triviality in R.

You can find all of my Advent of Code solutions on [GitHub](https://github.com/stephenfeagin/adventofcode).
