---
title: "Advent of Code: 2022 Day 2 in R"
date: 2023-09-07
categories:
  - r
  - advent of code
  - puzzle
format: gfm
---

### Read Input

As always, the first step is to read the input. I have saved the example data in [test_input.txt](test_input.txt) and the real puzzle input in [input.txt](input.txt). I know that the input has two letters per line separated by a space, which is exactly what `read.delim()` is made for. I'll specify `header = FALSE` and `sep = " "`, and keep the default variable names `V1` and `V2`.

```{r}
read_input <- function(file) {
  read.delim(file, header = FALSE, sep = " ")
}
```

Just to see what the test data looks like:
```{r}
test_data <- read_input("test_input.txt")
test_data
```

### Scoring Function

We need to create a function that will determine our total score based on what shape we played and what the outcome was. Remember for the first column, A is Rock, B is Paper, and C is Scissors. For the second column, X is Rock, Y is Paper, and Z is scissors. For the outcomes, you get 0 points for losing, 3 for drawing, and 6 for winning.

First, we have to determine the outcome. Unfortunately, this is somewhat tedious but the most straightforward way is to code it by hand.
```{r}
find_outcome <- function(p1, p2) {
  if (p1 == "A") {
    if (p2 == "X") return(3) # Rock-Rock=Draw
    if (p2 == "Y") return(6) # Rock-Paper=Win
    if (p2 == "Z") return(0) # Rock-Scissors=Lose
  }
  if (p1 == "B") {
    if (p2 == "X") return(0) # Paper-Rock=Lose
    if (p2 == "Y") return(3) # Paper-Paper=Draw
    if (p2 == "Z") return(6) # Paper-Scissors=Win
  }
  if (p1 == "C") {
    if (p2 == "X") return(6) # Scissors-Rock=Win
    if (p2 == "Y") return(0) # Scissors-Paper=Lose
    if (p2 == "Z") return(3) # Scissors-Scissors=Draw
  }
}
```

We can create a simple vector for points based on shape selection. You get 1 point for Rock (`"X"`), 2 points for Paper (`"Y"`), and 3 points for Scissors (`"Z"`). So I can make a vector of those shapes and pull out the scores using `which()`.

```{r}
find_shape_score <- function(shape) {
  shape_scores <- c("X", "Y", "Z")
  which(shape_scores == shape)
}
```

Finally, we need to combine those to get the total score.

```{r}
find_total_score <- function(p1, p2) {
  outcome <- find_outcome(p1, p2)
  shape_score <- find_shape_score(p2)
  outcome + shape_score
}
```

Let's apply that to the example data. I will add a column to the `test_data` data frame with the expected results from the example:

```{r}
test_data$expected <- c(8, 1, 6)
```

Then we can add a column with the calculated scores. Because I haven't written the functions to work with vectors of inputs, I iterate over the data frame.

```{r}
for (i in seq_len(nrow(test_data))) {
  test_data$calculated[i] <- find_total_score(test_data$V1[i], test_data$V2[i])
}
test_data
```
And in code, we can see:

```{r}
all(test_data$expected == test_data$calculated)
```


I'll just put that into a function to re-use for the real data:

```{r}
calculate_all_scores <- function(input_data) {
  calculated <- numeric(nrow(input_data))
  for (i in seq_len(nrow(input_data))) {
    calculated[i] <- find_total_score(input_data$V1[i], input_data$V2[i])
  }
  calculated
}
```

So the last part is getting the sum:

```{r}
part_1 <- function(input_data) {
  calculated <- calculate_all_scores(input_data)
  sum(calculated)
}
```

Try it with the test data:

```{r}
part_1(test_data)
```

Which is what we expect.

Now we try it on the real data:

```{r}
real_data <- read_input("input.txt")
part_1(real_data)
```

And that's correct!

## Part 2

```{r}
find_correct_shape <- function(p1, outcome) {
  if (p1 == "A") {
    if (outcome == "X") return("scissors") # Rock+Lose=Scissors
    if (outcome == "Y") return("rock")     # Rock+Draw=Rock
    if (outcome == "Z") return("paper")    # Rock+Win=Paper
  }
  if (p1 == "B") {
    if (outcome == "X") return("rock")     # Paper+Lose=Rock
    if (outcome == "Y") return("paper")    # Paper+Draw=Paper
    if (outcome == "Z") return("scissors") # Paper+Win=Scissors
  }
  if (p1 == "C") {
    if (outcome == "X") return("paper")    # Scissors+Lose=Paper
    if (outcome == "Y") return("scissors") # Scissors+Draw=Scissors
    if (outcome == "Z") return("rock")     # Scissors+Win=Rock
  }
}
```

I'll make a new `find_shape_score()` function using the actual shape names:
```{r}
find_shape_score <- function(shape) {
  shape_scores <- c("rock", "paper", "scissors")
  which(shape_scores == shape)
}
```

I also need to make a lookup for outcome points. I could write this into the bigger functions but I'll make it its own just to be super clear about what's going on.
```{r}
find_outcome_score <- function(outcome) {
  if (outcome == "X") return(0)
  if (outcome == "Y") return(3)
  if (outcome == "Z") return(6)
}
```

I need to make a new `find_total_score()` function as well:
```{r}
find_total_score <- function(p1, outcome) {
  correct_shape <- find_correct_shape(p1, outcome)
  shape_score <- find_shape_score(correct_shape)
  outcome_score <- find_outcome_score(outcome)
  shape_score + outcome_score
}
```

I'll test this out again on the test data. I read it in one more time just to get a clean slate since I added some extra columns in part 1. I can still use the same `calculate_all_scores()` function because I kept the same names for the intermediate functions, and that function just applies `find_total_score()` to the rows of the data frame.

```{r}
test_data <- read_input("test_input.txt")
test_data$expected <- c(4, 1, 7)
test_data$calculated <- calculate_all_scores(test_data)
test_data
```
Looks good! Now we just come up with the total:

```{r}
sum(test_data$calculated)
```
Which matches the example. Now for the real data:

```{r}
part_2 <- function(input_data) {
  calculated <- calculate_all_scores(input_data)
  sum(calculated)
}
part_2(real_data)
```

Which is right! There we go.
